---
title: "Dry run for scale-up simulation (no sampling)"
author: ""
#author: "Dennis Feehan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
#vignette: >
#  %\VignetteIndexEntry{Dry run for scale-up simulation (no sampling)}
#  %\VignetteEngine{knitr::rmarkdown}
#  \usepackage[utf8]{inputenc}
---


```{r, echo=FALSE, message=FALSE}
# don't include this vignette in pkgdown site
# see https://github.com/r-lib/pkgdown/issues/457
knitr::opts_chunk$set(eval = FALSE)

library(igraph)
library(plyr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(devtools)
library(gridExtra)
load_all()

knitr::opts_chunk$set(echo=FALSE, message=FALSE, fig.width=12, fig.height=12)

```


## Simulation study: dry run

This document is a dry run for the simulation study that is included as part
of the generalized scale-up paper (TODO LINK).

The plots below show the expected values of different estimators and parameters;
it does not conduct any sampling.

These plots show the expected results of the simulation
bias in two different scale-up estimators
under the two-parameter model when $\tau=1$ and $p_{F|H} = 1$.

```{r}

param.vals <- expand.grid(# size of total population, U
                          N=10e3,
                          # frame population is half of total
                          p.F=c(.1, .5, 1),
                          # hidden popn is 3 percent of total
                          p.H=.03,
                          # we'll assume all hidden popn members are in the frame popn
                          p.F.given.H=1,
                          # prob of edge between two nodes in the same group
                          zeta=.05,
                          # relative probability of edge between frame and not frame
                          xi=0.4,
                          # relative probability of edge between hidden and not hidden
                          rho=seq(from=.1, to=1, by=.1),
                          # true positive rate (1=accurate reports)
                          tau=c(.1, .5, 1))

all.param.list <- split(param.vals, rownames(param.vals))

## get a list of parameter objects based on the 1 param/nested specification
## now do the same for the two-param version
all.param.list.twoparam <- llply(all.param.list,
                                 sbm_params,
                                 type="4group_2param")

res.twoparam <- ldply(all.param.list.twoparam,
                      sbm_ev)
res.twoparam$type <- "twoparam"

res <- res.twoparam

res.nsum <- res %>% mutate(estimate=tau*d.F.H / dbar.U.F, 
                           estimator="nsum.est")

res.gnsum <- res %>% mutate(estimate=tau*d.F.H / (tau * dbar.H.F),
                            estimator="generalized.est")

res.both <- rbind(res.nsum, res.gnsum)
res.both <- res.both %>% mutate(N.H = N*p.H)
res.both <- res.both %>% mutate(bias = estimate - N.H)

## give estimators nice labels
res.both <- res.both %>% 
            mutate(estimator.label=ifelse(estimator=="generalized", 
                                                     "generalized scale-up",
                                                     "basic scale-up"))

```

```{r, results='asis', fig.width=6, fig.height=10}

toplot.bias.twoparam <- res.both

#toplot.bias.twoparam <- res.both %>% 
#                  filter(
#                         # NA correspondes to p_{F|H} = p_F
#                         #is.na(p.F.given.H),
#                         dbl.in(p.F.given.H, 1),
#                         #dbl.in(tau, res.vals),
#                         dbl.in(tau, 1),
#                         #dbl.in(rho, res.vals),
#                         dbl.in(p.F, res.vals)
#                         #dbl.in(xi, 0.6),
#                         #dbl.in(xi, res.vals)
#                         ) %>%
#                  filter(type == "twoparam")

#bias.twoparam <- ggplot(toplot.bias.twoparam) +
#                  geom_line(aes(x=rho, y=abs(bias), 
#                                color=estimator, 
#                                group=estimator)) +
#                  geom_hline(aes(yintercept=1), linetype=1, color="gray") +
#                  facet_grid(tau  ~ p.F,
#                             labeller = labeller(tau=label_bquote(tau == .(tau)),
#                                                 p.F=label_bquote(p[F] == .(p.F)))) +
#                  #facet_grid(xi  ~ p.F,
#                  #           labeller = labeller(xi=label_bquote(xi == .(xi)),
#                  #                               p.F=label_bquote(p[F] == .(p.F)))) +
#                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
#                  ylab(bquote(paste("|bias|"))) +
#                  theme(legend.position="bottom",
#                        axis.text.x=element_text(angle=90, vjust=0.5),
#                        panel.border=element_rect(size=.5, fill=NA),
#                        panel.background=element_blank(),
#                        panel.grid.major=element_blank(),
#                        panel.grid.minor=element_blank()) +
#                  ggtitle(bquote(paste(tau, "=1, ", paste(p,'[F|H]'), "= 1")))

#bias.twoparam

```



