---
title: "Explore expected values for stochastic block-model simulations"
author: ""
#author: "Dennis Feehan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sbm-simulation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE, message=FALSE}
library(igraph)
library(plyr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(devtools)
load_all()

knitr::opts_chunk$set(echo=FALSE, message=FALSE)

```

## Overview

This file shows the (analytically-derived) expected values for several
quantities under the stochastic block model simulation design. There are
three variants of the design: simple, nested, and two-parameter. 
These can be distinguished by the mixing matrices, as shown below.

#### Simple mixing matrix

In the simple mixing matrix, the probability of a tie between any two
vertices that are not in exactly the same group is multiplied by $\rho$:

$$
\begin{matrix}
    & ~F~H           & ~F \lnot H     & \lnot F ~H     & \lnot F \lnot H \\
    ~F~H            & \zeta           & \rho\cdot\zeta  & \rho\cdot\zeta  & \rho\cdot\zeta \\
    ~F \lnot H      & \rho\cdot\zeta  & \zeta           & \rho\cdot\zeta  & \rho\cdot\zeta \\
    \lnot F ~H      & \rho\cdot\zeta  & \rho\cdot\zeta  & \zeta           & \rho\cdot\zeta \\
    \lnot F \lnot H & \rho\cdot\zeta  & \rho\cdot\zeta  & \rho\cdot\zeta  & \zeta
\end{matrix}
$$

#### Nested mixing matrix

In the nested mixing matrix, the probability of a tie between any two vertices
who differ in one dimension (membership in $F$ or membership in $H$) is multiplied by
$\rho$; the probability of a tie between two vertices that differ on both dimensions
is multiplied by $\rho^2$:

$$
\begin{matrix}
    & ~F~H           & ~F \lnot H     & \lnot F ~H     & \lnot F \lnot H \\
    ~F~H            & \zeta           & \rho\cdot\zeta  & \rho\cdot\zeta  & \rho^2\cdot\zeta \\
    ~F \lnot H      & \rho\cdot\zeta  & \zeta           & \rho^2\cdot\zeta& \rho\cdot\zeta \\
    \lnot F ~H      & \rho\cdot\zeta  & \rho^2\cdot\zeta& \zeta           & \rho\cdot\zeta \\
    \lnot F \lnot H & \rho^2\cdot\zeta& \rho\cdot\zeta  & \rho\cdot\zeta  & \zeta
\end{matrix}
$$

#### Two-parameter mixing matrix

In the two-parameter mixing matrix, the probability of a tie between two vertices
that differ in membership in $F$ is multiplied by $\xi$; and, independently, the
probability of a tie between two vertices that differ in membership in $H$ is
multiplied by $\rho$:

$$
\begin{matrix}
                    & ~F~H                   & ~F \lnot H              & \lnot F ~H              & \lnot F \lnot H \\
    ~F~H            & \zeta                  & \rho\cdot\zeta          & \xi\cdot\zeta           & \xi\cdot\rho\cdot\zeta \\
    ~F \lnot H      & \rho\cdot\zeta         & \zeta                   & \xi\cdot\rho\cdot\zeta  & \xi\cdot\zeta \\
    \lnot F ~H      & \xi\cdot\zeta          & \xi\cdot\rho\cdot\zeta  & \zeta                   & \rho\cdot\zeta \\
    \lnot F \lnot H & \xi\cdot\rho\cdot\zeta & \xi\cdot\zeta           & \rho\cdot\zeta          & \zeta
\end{matrix}
$$

### Analytical derivations

The expected values for the number of connections between groups or sets of
groups depends on the structure of the mixing matrix.
Write the number of vertices in each of the four blocks as
a vector, $\mathbf{n}$.

Suppose we wish to count the number of expected connections from
a set $A$ to another set $B$, with both $A$ and $B$ built up from
one or more of the four blocks. Let $\mathbf{a}$ be the 0/1 vector
that indicates which blocks $A$ contains, and $\mathbf{b}$ be the 0/1
vector that indicates which blocks $B$ contains. The size of A
is then $N_A = \mathbf{a}^T \mathbf{n}$.

Let $\mathbf{a}_N = \mathbf{a} * \mathbf{n}$, where $*$
denotes element-wise multiplication. Then

$$
\begin{equation}
d_{A,B} = \mathbf{a}_N^T ~M~ \mathbf{b}_N - 
          (\mathbf{a} * \mathbf{b} * \mathbf{n})^T \text{diag}(M).
\end{equation}
$$

To understand this expression, note that the number of edges
between two distinct blocks $i$ and $j$ will be
$N_i~N_j~M_{ij}$. Since we do not permit self-loops, 
the number of edges from one block to itself
will be $\frac{N_i~(N_i - 1)}{2}~M_{ii}$.  The expression

$$
\begin{equation}
\mathbf{a}_N^T~M~\mathbf{b}_N = \sum_{i \in b(A)} \sum_{j \in b(B)} N_i~N_j~M_{ij}
\end{equation}
$$
where $b(A)$ and $b(B)$ are the sets of blocks consisting of $A$ and $B$ is
almost what we want. The problem is that each block $i$ in both $b(A)$ and
$b(B)$ is sending edges to itself. In the equation above, it contributes
$N_i^2~M_{i,i}$ instead of $\frac{N_i~(N_i - 1)}{2}~M_{i,i}$. So instead of the equation
above, we actually want

$$
\begin{equation}
d_{A, B} = \sum_{i \in b(A)} \sum_{j \in b(B)} N_i~N_j~M_{ij} -
   \sum_{i \in b(A) \cap b(B)} \frac{(N_i^2 + N_i)}{2}~M_{i,i},
\end{equation}
$$

In this equation, the subtracted term accounts for the difference between
$N_i^2$ and $\frac{N_i~(N_i - 1)}{2}$.

### TODO - also show expression for variance of expected values



## Illustration

Next, we will look at the expected values for several important quantities under both versions
of the stochastic block-model: the nested and the simple versions.

```{r}
tmp.params <- list(N=100,
                   p.F=.5,
                   p.H=1,
                   pi.within=.1,
                   rho=1,
                   tau=1)

params <- sbm_params(tmp.params, type="4group_1param_simple")

tmp.ev <- sbm_ev(params)

tmp.block.sizes <- params$block.sizes
this.pref.matrix <- params$pref.matrix

```

```{r}

param.vals <- expand.grid(# size of total population, U
                          N=10e3,
                          #N=10e3,
                          # frame population is half of total
                          #p.F=c(.1,.5,1),
                          p.F=seq(from=.1, to=1, by=.1),
                          # hidden popn is 3 percent of total
                          p.H=.03,
                          # prob of edge between two nodes in the same group
                          pi.within=.05,
                          # relative prob of edge between two nodes not in the same group
                          #rho=c(.1, .5, 1),
                          rho=seq(from=.1, to=1, by=.1),
                          #tau=c(.1, .5, 1))
                          tau=seq(from=.1, to=1, by=.1))

all.param.list <- split(param.vals, rownames(param.vals))

```

```{r}
all.param.list.nested <- llply(all.param.list,
                               sbm_params,
                               type="4group_1param_nested")

res.nested <- ldply(all.param.list.nested,
                    sbm_ev)

res.nested$type <- "nested"

## now do the same for the simple version

all.param.list.simple <- llply(all.param.list,
                               sbm_params,
                               type="4group_1param_simple")

res.simple <- ldply(all.param.list.simple,
                    sbm_ev)
res.simple$type <- "simple"

res <- rbind(res.nested, res.simple)

#res.nsum <- res %>% mutate(estimate=N*p.H*total, 
res.nsum <- res %>% mutate(estimate=tau*d.F.H / dbar.U.F, 
                           estimator="nsum.est")

#res.gnsum <- res %>% mutate(estimate=N*p.H, 
res.gnsum <- res %>% mutate(estimate=tau*d.F.H / (tau * dbar.H.F),
                            estimator="generalized.est")

res.both <- rbind(res.nsum, res.gnsum)
res.both <- res.both %>% mutate(N.H = N*p.H)

```

```{r}

label_equals <- function(variable, value) {

    ## turn, eg p.F into p[F]
    variable <- str_replace(variable, "\\.(\\w+)", "[\\1]")

    #llply(as.character(variable), function(x) parse(text=paste(x, "==", v))
    #llply(as.character(variable), 
    llply(1:length(value),
          function(idx) {
              x <- as.character(variable)
              v <- as.character(value[idx])
              parse(text=paste(x, "==", v))
          })
}

## NOTE: this function comes from the R Cookbook site
##
## http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
##
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
    require(grid)

    # Make a list from the ... arguments and plotlist
    plots <- c(list(...), plotlist)

    numPlots = length(plots)

    # If layout is NULL, then use 'cols' to determine layout
    if (is.null(layout)) {
        # Make the panel
        # ncol: Number of columns of plots
        # nrow: Number of rows needed, calculated from # of cols
        layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                         ncol = cols, nrow = ceiling(numPlots/cols))
    }

    if (numPlots==1) {
        print(plots[[1]])

    } else {
        # Set up the page
        grid.newpage()
        pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

        # Make each plot, in the correct location
        for (i in 1:numPlots) {
            # Get the i,j matrix positions of the regions that contain this subplot
            matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

            print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                            layout.pos.col = matchidx$col))
        }
    }
}
```

```{r}
res.vals <- c(.1, .5, 1)
toplot<- res.both %>% 
         filter(tau %in% res.vals &
                #rho %in% res.vals &
                p.F %in% res.vals)

```

### Expected values for adjustment factors

```{r, fig.width=14, fig.height=6, results='asis'}

phi.byrho.lim <- ggplot(toplot %>%
                             mutate(estimator=ifelse(estimator=="generalized.est",
                                                     "generalized scale-up",
                                                     "basic scale-up"))) +
                  geom_line(aes(x=rho, y=phi.F, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=1), linetype=1, color="gray") +
                  facet_grid(tau ~ p.F,
                             labeller = label_equals) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank())

delta.byrho.lim <- ggplot(toplot %>%
                             mutate(estimator=ifelse(estimator=="generalized.est",
                                                     "generalized scale-up",
                                                     "basic scale-up"))) +
                  geom_line(aes(x=rho, y=delta.F, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=1), linetype=1, color="gray") +
                  facet_grid(tau ~ p.F,
                             labeller = label_equals) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank())

phidelta.byrho.lim <- ggplot(toplot %>%
                             mutate(estimator=ifelse(estimator=="generalized.est",
                                                     "generalized scale-up",
                                                     "basic scale-up"))) +
                  geom_line(aes(x=rho, y=deltaXphi.F, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=1), linetype=1, color="gray") +
                  facet_grid(tau ~ p.F,
                             labeller = label_equals) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank())

multiplot(phi.byrho.lim, delta.byrho.lim, phidelta.byrho.lim, cols=3)
```

### Basic scale-up estimator expected values 

```{r, fig.height=6, fig.width=6}

ev.byrho.lim <- ggplot(toplot %>%
                       filter(estimator != "generalized.est") %>%
                       mutate(estimator=ifelse(estimator=="generalized.est",
                                                          "generalized scale-up",
                                                          "basic scale-up"))) +
                  geom_line(aes(x=rho, y=estimate, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=N.H), linetype=1, color="gray") +
                  scale_shape_manual(values=c("generalized scale-up"=4,
                                              "basic scale-up"=3)) + 
                  facet_grid(tau ~ p.F,
                             labeller = label_equals) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank())

ev.byrho.lim

```


```{r}

##ggsave(file=file.path(out.dir, "sim-expectedvalue-byrho-lim.pdf"),
##       plot=ev.byrho.lim,
##       width=5, height=5.5)
                
```

