---
title: "Explore expected values for stochastic block-model simulations"
author: ""
#author: "Dennis Feehan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sbm-simulation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE, message=FALSE}
library(igraph)
library(plyr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(devtools)
library(gridExtra)
load_all()

knitr::opts_chunk$set(echo=FALSE, message=FALSE, fig.width=12, fig.height=12)

```

## Overview

This file shows the (analytically-derived) expected values for several
quantities under the stochastic block model simulation design. There are
three variants of the design: simple, nested, and two-parameter. 
These can be distinguished by the mixing matrices, as shown below.

#### Simple mixing matrix

In the simple mixing matrix, the probability of a tie between any two
vertices that are not in exactly the same group is multiplied by $\rho$:

$$
\begin{matrix}
    & ~F~H           & ~F \lnot H     & \lnot F ~H     & \lnot F \lnot H \\
    ~F~H            & \zeta           & \rho\cdot\zeta  & \rho\cdot\zeta  & \rho\cdot\zeta \\
    ~F \lnot H      & \rho\cdot\zeta  & \zeta           & \rho\cdot\zeta  & \rho\cdot\zeta \\
    \lnot F ~H      & \rho\cdot\zeta  & \rho\cdot\zeta  & \zeta           & \rho\cdot\zeta \\
    \lnot F \lnot H & \rho\cdot\zeta  & \rho\cdot\zeta  & \rho\cdot\zeta  & \zeta
\end{matrix}
$$

#### Nested mixing matrix

In the nested mixing matrix, the probability of a tie between any two vertices
who differ in one dimension (membership in $F$ or membership in $H$) is multiplied by
$\rho$; the probability of a tie between two vertices that differ on both dimensions
is multiplied by $\rho^2$:

$$
\begin{matrix}
    & ~F~H           & ~F \lnot H     & \lnot F ~H     & \lnot F \lnot H \\
    ~F~H            & \zeta           & \rho\cdot\zeta  & \rho\cdot\zeta  & \rho^2\cdot\zeta \\
    ~F \lnot H      & \rho\cdot\zeta  & \zeta           & \rho^2\cdot\zeta& \rho\cdot\zeta \\
    \lnot F ~H      & \rho\cdot\zeta  & \rho^2\cdot\zeta& \zeta           & \rho\cdot\zeta \\
    \lnot F \lnot H & \rho^2\cdot\zeta& \rho\cdot\zeta  & \rho\cdot\zeta  & \zeta
\end{matrix}
$$

#### Two-parameter mixing matrix

In the two-parameter mixing matrix, the probability of a tie between two vertices
that differ in membership in $F$ is multiplied by $\xi$; and, independently, the
probability of a tie between two vertices that differ in membership in $H$ is
multiplied by $\rho$:

$$
\begin{matrix}
                    & ~F~H                   & ~F \lnot H              & \lnot F ~H              & \lnot F \lnot H \\
    ~F~H            & \zeta                  & \rho\cdot\zeta          & \xi\cdot\zeta           & \xi\cdot\rho\cdot\zeta \\
    ~F \lnot H      & \rho\cdot\zeta         & \zeta                   & \xi\cdot\rho\cdot\zeta  & \xi\cdot\zeta \\
    \lnot F ~H      & \xi\cdot\zeta          & \xi\cdot\rho\cdot\zeta  & \zeta                   & \rho\cdot\zeta \\
    \lnot F \lnot H & \xi\cdot\rho\cdot\zeta & \xi\cdot\zeta           & \rho\cdot\zeta          & \zeta
\end{matrix}
$$

### Analytical derivations

The expected values for the number of connections between groups or sets of
groups depends on the structure of the mixing matrix.
Write the number of vertices in each of the four blocks as
a vector, $\mathbf{n}$.

Suppose we wish to count the number of expected connections from
a set $A$ to another set $B$, with both $A$ and $B$ built up from
one or more of the four blocks. Let $\mathbf{a}$ be the 0/1 vector
that indicates which blocks $A$ contains, and $\mathbf{b}$ be the 0/1
vector that indicates which blocks $B$ contains. The size of A
is then $N_A = \mathbf{a}^T \mathbf{n}$.

Let $\mathbf{a}_N = \mathbf{a} * \mathbf{n}$, where $*$
denotes element-wise multiplication. Then

$$
\begin{equation}
d_{A,B} = \mathbf{a}_N^T ~M~ \mathbf{b}_N - 
          (\mathbf{a} * \mathbf{b} * \mathbf{n})^T \text{diag}(M).
\end{equation}
$$

To understand this expression, note that the number of edges
between two distinct blocks $i$ and $j$ will be
$N_i~N_j~M_{ij}$. Since we do not permit self-loops, 
the number of edges from one block to itself
will be $\frac{N_i~(N_i - 1)}{2}~M_{ii}$.  The expression

$$
\begin{equation}
\mathbf{a}_N^T~M~\mathbf{b}_N = \sum_{i \in b(A)} \sum_{j \in b(B)} N_i~N_j~M_{ij}
\end{equation}
$$
where $b(A)$ and $b(B)$ are the sets of blocks consisting of $A$ and $B$ is
almost what we want. The problem is that each block $i$ in both $b(A)$ and
$b(B)$ is sending edges to itself. In the equation above, it contributes
$N_i^2~M_{i,i}$ instead of $\frac{N_i~(N_i - 1)}{2}~M_{i,i}$. So instead of the equation
above, we actually want

$$
\begin{equation}
d_{A, B} = \sum_{i \in b(A)} \sum_{j \in b(B)} N_i~N_j~M_{ij} -
   \sum_{i \in b(A) \cap b(B)} \frac{(N_i^2 + N_i)}{2}~M_{i,i},
\end{equation}
$$

In this equation, the subtracted term accounts for the difference between
$N_i^2$ and $\frac{N_i~(N_i - 1)}{2}$.

<!--TODO - also show expression for variance of expected values-->

## Illustration

Next, we will look at the expected values for several important quantities
under all of the versions of the stochastic block-model.

```{r}

step.size <- .2

param.vals <- expand.grid(# size of total population, U
                          N=10e3,
                          # frame population is half of total
                          p.F=seq(from=.2, to=1, by=step.size),
                          # hidden popn is 3 percent of total
                          p.H=.03,
                          p.F.given.H=1,
                          # prob of edge between two nodes in the same group
                          zeta=.05,
                          xi=seq(from=.2, to=1, by=step.size),
                          rho=seq(from=.2, to=1, by=step.size),
                          tau=seq(from=.2, to=1, by=step.size))

all.param.list <- split(param.vals, rownames(param.vals))

```

```{r}

## get a list of parameter objects based on the 1 param/nested specification
all.param.list.nested <- llply(all.param.list,
                               sbm_params,
                               type="4group_1param_nested")

res.nested <- ldply(all.param.list.nested,
                    sbm_ev)

res.nested$type <- "nested"

## now do the same for the simple version
all.param.list.simple <- llply(all.param.list,
                               sbm_params,
                               type="4group_1param_simple")

res.simple <- ldply(all.param.list.simple,
                    sbm_ev)
res.simple$type <- "simple"

## now do the same for the two-param version
all.param.list.twoparam <- llply(all.param.list,
                                 sbm_params,
                                 type="4group_2param")

res.twoparam <- ldply(all.param.list.twoparam,
                      sbm_ev)
res.twoparam$type <- "twoparam"


res <- rbind(res.nested, res.simple, res.twoparam)

res.nsum <- res %>% mutate(estimate=tau*d.F.H / dbar.U.F, 
                           estimator="nsum.est")

res.ansum <- res %>% mutate(estimate=tau*d.F.H / dbar.F.F,
                            estimator="adapted.est")

res.gnsum <- res %>% mutate(estimate=tau*d.F.H / (tau * dbar.H.F),
                            estimator="generalized.est")

res.both <- rbind(res.nsum, res.ansum, res.gnsum)
res.both <- res.both %>% mutate(N.H = N*p.H)
res.both <- res.both %>% mutate(bias = N.H - estimate)

## give estimators nice labels
res.both <- res.both %>% 
            mutate(estimator.label=ifelse(estimator=="generalized", 
                                                     "generalized scale-up",
                                                     ifelse(estimator=="adapted", 
                                                            "adapted scale-up", 
                                                            "basic scale-up")))

```

```{r}

res.vals <- c(.2, .6, 1)

dbl.in <- function(a, vals, tol=1e-10) {
    laply(a, 
          function(x) {
            return(any(abs(x-vals) < tol))
          })
}

xi.fixed <- 0.6
toplot.fix.xi <- res.both %>% 
                 filter(dbl.in(tau, res.vals),
                        dbl.in(p.F, res.vals),
                        dbl.in(xi, xi.fixed)) %>%
                 mutate(type=ifelse(type=="twoparam",
                                    paste0("twoparam (xi=", xi.fixed, ")"),
                                    type))

rho.fixed <- 0.6
toplot.fix.rho <- res.both %>% 
                 filter(dbl.in(tau, res.vals),
                        dbl.in(p.F, res.vals),
                        dbl.in(rho, rho.fixed)) %>%
                 mutate(type=ifelse(type=="twoparam",
                                    paste0("twoparam (rho=", rho.fixed, ")"),
                                    type))

```

### Expected values for adjustment factors

First we'll look at these holding $\xi$ fixed at `r xi.fixed` (which only affects the twoparam model):

```{r, fig.width=6, fig.height=15, results='asis'}

phi.byrho.lim <- ggplot(toplot.fix.xi) +
                  geom_line(aes(x=rho, y=phi.F, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=1), linetype=1, color="gray") +
                  facet_grid(tau ~ p.F,
                            labeller = labeller(rows=label_bquote(tau == .(tau)),
                                                cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
                  ylab(bquote(paste("frame ratio (", phi[F], ")"))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank())

delta.byrho.lim <- ggplot(toplot.fix.xi) +
                  geom_line(aes(x=rho, y=delta.F, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=1), linetype=1, color="gray") +
                  facet_grid(tau ~ p.F,
                             labeller = labeller(rows=label_bquote(tau == .(tau)),
                                                 cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
                  ylab(bquote(paste("degree ratio (", delta[F], ")"))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank())

phidelta.byrho.lim <- ggplot(toplot.fix.xi) +
                  geom_line(aes(x=rho, y=deltaXphi.F, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=1), linetype=1, color="gray") +
                  facet_grid(tau ~ p.F,
                             labeller = labeller(rows=label_bquote(tau == .(tau)),
                                                 cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
                  ylab(bquote(paste("partial adjustment factor (", delta[F]~phi[F], ")"))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank())

grid.arrange(phi.byrho.lim, 
             delta.byrho.lim, 
             phidelta.byrho.lim, 
             nrow=3)
```

Next we'll look at these holding $\rho$ fixed at `r rho.fixed`
(which only affects the twoparam model):

```{r, fig.width=6, fig.height=15, results='asis'}

phi.byxi.lim <- ggplot(toplot.fix.rho) +
                  geom_line(aes(x=xi, y=phi.F, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=1), linetype=1, color="gray") +
                  facet_grid(tau ~ p.F,
                             labeller = labeller(rows=label_bquote(tau == .(tau)),
                                                 cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(xi))) +
                  ylab(bquote(paste("frame ratio (", phi[F], ")"))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank())

delta.byxi.lim <- ggplot(toplot.fix.rho) +
                  geom_line(aes(x=xi, y=delta.F, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=1), linetype=1, color="gray") +
                  facet_grid(tau ~ p.F,
                             labeller = labeller(rows=label_bquote(tau == .(tau)),
                                                 cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(xi))) +
                  ylab(bquote(paste("degree ratio (", delta[F], ")"))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank())

phidelta.byxi.lim <- ggplot(toplot.fix.rho) +
                  geom_line(aes(x=xi, y=deltaXphi.F, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=1), linetype=1, color="gray") +
                  facet_grid(tau ~ p.F,
                             labeller = labeller(rows=label_bquote(tau == .(tau)),
                                                 cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(xi))) +
                  ylab(bquote(paste("partial adjustment factor (", delta[F]~phi[F], ")"))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank())

grid.arrange(phi.byxi.lim, 
             delta.byxi.lim, 
             phidelta.byxi.lim, 
             nrow=3)
```

### Bias

#### Two parameter model

These plots show the bias in three different scale-up estimators
under the two-parameter model when $\tau=1$.

```{r, results='asis', fig.width=6, fig.height=10}

toplot.bias.twoparam <- res.both %>% 
                  filter(
                         #dbl.in(tau, res.vals),
                         dbl.in(tau, 1),
                         #dbl.in(rho, res.vals),
                         dbl.in(p.F, res.vals)
                         #dbl.in(xi, 0.6),
                         #dbl.in(xi, res.vals)
                         ) %>%
                  filter(type == "twoparam")

bias.twoparam <- ggplot(toplot.bias.twoparam) +
                  #geom_line(aes(x=rho, y=bias, color=xi, group=xi)) +
                  geom_line(aes(x=rho, y=abs(bias), color=estimator)) +
                  geom_hline(aes(yintercept=1), linetype=1, color="gray") +
                  facet_grid(xi  ~ p.F,
                             labeller = labeller(rows=label_bquote(xi == .(xi)),
                                                 cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
                  ylab(bquote(paste("|bias|"))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank()) +
                  ggtitle(bquote(paste(tau, "=1")))

bias.twoparam

```

```{r}
## helper fns for plots

estimator_shape_scale <- scale_shape_manual(values=c("generalized scale-up"=5, 
                                                     "adapted scale-up"=6, 
                                                     "basic scale-up"=1))



```

### Basic scale-up estimator expected values

```{r, fig.height=6, fig.width=6, res='asis'}

bnsum.ev.byrho.lim <- ggplot(toplot.fix.xi %>%
                       filter(estimator == "nsum.est")) +
                  geom_line(aes(x=rho, y=estimate, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=N.H), linetype=1, color="gray") +
                  estimator_shape_scale +
                  facet_grid(tau ~ p.F,
                             labeller = labeller(rows=label_bquote(tau == .(tau)),
                                                 cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank()) +
                  ggtitle("Expected value: basic scale-up")

bnsum.ev.byrho.lim

```

### Adapted scale-up estimator expected values 

```{r, fig.height=6, fig.width=6, results='asis'}

ansum.ev.byrho.lim <- ggplot(toplot.fix.xi %>%
                       filter(estimator == "adapted.est")) +
                  geom_line(aes(x=rho, y=estimate, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=N.H), linetype=1, color="gray") +
                  estimator_shape_scale +
                  facet_grid(tau ~ p.F,
                             labeller = labeller(rows=label_bquote(tau == .(tau)),
                                                 cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank()) +
                  ggtitle("Expected value: adapted scale-up")

ansum.ev.byrho.lim

```


### Generalized scale-up estimator expected values 

```{r, fig.height=6, fig.width=6, results='asis'}

gnsum.ev.byrho.lim <- ggplot(toplot.fix.xi %>%
                       filter(estimator == "generalized.est")) +
                  geom_line(aes(x=rho, y=estimate, color=type, linetype=type)) +
                  geom_hline(aes(yintercept=N.H), linetype=1, color="gray") +
                  estimator_shape_scale +
                  facet_grid(tau ~ p.F,
                             labeller = labeller(rows=label_bquote(tau == .(tau)),
                                                 cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~(rho))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank()) +
                  ggtitle("Expected value: generalized scale-up")

gnsum.ev.byrho.lim

```


## Scenario: Two-parameter model

Now we look at a specific scenario for just the two-parameter model.

```{r}

step.size <- .1

param.vals <- expand.grid(# size of total population, U
                          N=10e2,
                          # frame population is half of total
                          p.F=seq(from=.2, to=1, by=step.size),
                          # hidden popn is 3 percent of total
                          p.H=.03,
                          p.F.given.H=c(0, .5, 1, NA),
                          # prob of edge between two nodes in the same group
                          zeta=.05,
                          xi=seq(from=.2, to=1, by=step.size),
                          rho=seq(from=.2, to=1, by=step.size),
                          #rho=.6,
                          tau=1)
                          #tau=seq(from=.2, to=1, by=step.size))

all.param.list <- split(param.vals, rownames(param.vals))

```

```{r}

## get a list of parameter objects based on the 1 param/nested specification
## now do the same for the two-param version
all.param.list.twoparam <- llply(all.param.list,
                                 sbm_params,
                                 type="4group_2param")

res.twoparam <- ldply(all.param.list.twoparam,
                      sbm_ev)
res.twoparam$type <- "twoparam"

res <- res.twoparam

res.nsum <- res %>% mutate(estimate=tau*d.F.H / dbar.U.F, 
                           estimator="nsum.est")

res.ansum <- res %>% mutate(estimate=tau*d.F.H / dbar.F.F,
                            estimator="adapted.est")

res.gnsum <- res %>% mutate(estimate=tau*d.F.H / (tau * dbar.H.F),
                            estimator="generalized.est")

res.both <- rbind(res.nsum, res.ansum, res.gnsum)
res.both <- res.both %>% mutate(N.H = N*p.H)
res.both <- res.both %>% mutate(bias = N.H - estimate)

## give estimators nice labels
res.both <- res.both %>% 
            mutate(estimator.label=ifelse(estimator=="generalized", 
                                                     "generalized scale-up",
                                                     ifelse(estimator=="adapted", 
                                                            "adapted scale-up", 
                                                            "basic scale-up")))

```

```{r, results='asis', fig.width=12, fig.height=12}

tol <- 1e-2

#############
## fixing rho at 1
res.fixrho <- res.both %>% filter(rho==1,
                                  tau==1)
bias.twoparam <- ggplot(res.fixrho) +
                  geom_line(aes(x=xi, y=abs(bias), color=estimator)) +
                  facet_grid(p.F.given.H  ~ p.F,
                             labeller = labeller(rows=label_bquote(tau == .(tau)),
                                                 cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~F~and~not~F~(xi))) +
                  ylab(bquote(paste("|bias|"))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank()) +
                  ggtitle(bquote(paste(rho, "=1 and ", tau, "=1")))

bias.twoparam

#############
## fixing xi at 0.6
res.fixxi <- res.both %>% filter(dbl.in(xi, 0.6),
                                  tau==1) %>%
             filter(estimator != 'adapted.est')

bias.twoparam <- ggplot(res.fixxi) +
                  geom_line(aes(x=rho, y=abs(bias), color=estimator)) +
                  facet_grid(p.F.given.H  ~ p.F,
                             labeller = labeller(rows=label_bquote(tau == .(tau)),
                                                 cols=label_bquote(p[F] == .(p.F)))) +
                  xlab(bquote(amount~of~inhomogenous~mixing~F~and~not~F~(xi))) +
                  ylab(bquote(paste("|bias|"))) +
                  theme(legend.position="bottom",
                        axis.text.x=element_text(angle=90, vjust=0.5),
                        panel.border=element_rect(size=.5, fill=NA),
                        panel.background=element_blank(),
                        panel.grid.major=element_blank(),
                        panel.grid.minor=element_blank()) +
                  ggtitle(bquote(paste(xi, "=0.6 and ", tau, "=1")))

bias.twoparam

```

