---
title: "Scratch"
author: "Dennis Feehan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
#vignette: >
#  %\VignetteIndexEntry{Scratch}
#  %\VignetteEngine{knitr::rmarkdown}
#  \usepackage[utf8]{inputenc}
---


```{r}
library(knitr)
opts_chunk$set(eval=FALSE)
# don't include this vignette in pkgdown site
# see https://github.com/r-lib/pkgdown/issues/457
knitr::opts_chunk$set(eval = FALSE)

```

```{r}
library(igraph)
library(dplyr)
library(tidyr)
library(stringr)
```

```{r}
set.seed(12345)
```

Now set up the params for one random draw.

```{r}
num.sims <- 2

## TODO - parameterize by avg degree w/in group or something?

sim.params <- expand.grid(# size of total population, U
                          N=10e3,
                          #N=10e3,
                          # frame population is half of total
                          p.F=.5,
                          # hidden popn is 3 percent of total
                          p.H=.03,
                          # prob of edge between two nodes in the same group
                          pi.within=.05,
                          # relative prob of edge between two nodes not in the same group
                          rho=1,
                          tau=.9)


```

Now draw the actual random graph.

TODO - LEFT OFF HERE

* converting sim code to package
* I think this test file has slightly old version of code
* code up sim using the package (specify version)
* re-run sim to specify according to writeup, if necessary

TODO - TO DEMONSTRATE

* report.sbm.edges

TODO - UNIT TESTS

* report.sbm.edges (be sure to mention counter-intuitive in vs out arg)
* reporting.estimates

etc...

```{r}
this.g <- draw.4group.graph(sim.params, type="nested")
this.g.df <- get.data.frame(this.g, 'vertices')
```

Next, generate the reporting graph.

```{r}
this.r <- reporting.graph(this.g, 
                          hidden.popn=c('FH', 'notFH'),
                          frame.popn=c('FH', 'FnotH'))
```

Finally, produce estimates

```{r}

these.ests <- reporting.estimates(this.r, 
                                  hidden.popn=c('FH', 'notFH'),
                                  frame.popn=c('FH', 'FnotH'))

```

```{r}

#system.time(
#this.g <- draw.4group.graph(sim.params)
#)

#system.time(
#this.r <- reporting.graph(this.g, 
#                          hidden.popn=c('FH', 'notFH'),
#                          frame.popn=c('FH', 'FnotH'))
#)

this.r.df <- get.data.frame(this.r, 'vertices')

this.r.df %>% group_by(group) %>% summarise(diff=mean(d.degree-y.degree),d.bar=mean(d.degree))

this.r.df %>% group_by(group) %>% summarise(v.bar=mean(v.degree))


```

## Code checks

TODO - these should be made into unit tests...

Here are a few internal consistency checks that I ran to be sure the
simulation code is working.

This should be TRUE:
```{r}
this.g.df <- get.data.frame(this.g, 'vertices')

# if p.F = 1, then this 
if (! "d.notFnotH" %in% colnames(this.g.df)) {
    this.g.df$d.notFnotH <- 0
}

with(this.g.df, all(d.degree == d.FnotH + d.FH + d.notFnotH + d.notFH))
with(this.g.df, all(d.degree == d.FnotH + d.FH + d.notFH))
```

The `diff` column of the dataframe produced here should all be zeroes.
```{r}
tmp1 <- this.g.df %>% select(starts_with("d."), -contains("degree")) %>% 
  gather(group, edgecount) %>%
  group_by(group) %>% summarise(edgetot = sum(edgecount)) %>%
  mutate(group = str_sub(group, 3))
tmp2 <- this.g.df %>% group_by(group) %>% summarise(d.tot=sum(d.degree), n=n())
tmp <- tmp1 %>% left_join(tmp2) %>% mutate(diff = edgetot - d.tot)
tmp
```

This should be TRUE
```{r}
sum(tmp$edgetot)==sum(tmp$d.tot)
```

The `diff` column of the dataframe produced here should all be zeroes.
```{r}
this.r.df <- get.data.frame(this.r, 'vertices')

tmp1 <- this.r.df %>% select(starts_with("y."), -contains("degree")) %>%
        gather(group, edgecount) %>%
        group_by(group) %>% summarise(edgetot=sum(edgecount))  %>%
        mutate(group = str_sub(group, 3))

tmp2 <- this.r.df %>% group_by(group) %>% 
        summarise(y.tot=sum(y.degree), n=n())
tmp <- tmp1 %>% left_join(tmp2) %>% mutate(diff = edgetot - y.tot)
tmp

```

TODO - check tau

```{r}
sim.params <- expand.grid(# size of total population, U
                          N=10e1,
                          #N=10e3,
                          # frame population is half of total
                          p.F=c(0.5),
                          # hidden popn is 3 percent of total
                          p.H=.5,
                          # prob of edge between two nodes in the same group
                          pi.within=1,
                          # relative prob of edge between two nodes not in the same group
                          rho=c(.5),
                          tau=c(.5))


set.seed(100)

system.time(
sim.graph.g <- draw.4group.graph(sim.params)
)

system.time(
sim.graph.r <- reporting.graph(sim.graph.g, 
                          hidden.popn=c('FH', 'notFH'),
                          frame.popn=c('FH', 'FnotH'))
)

h.idx <- V(sim.graph.r)[group %in% c('FH', 'notFH')] 
h1.idx <- V(sim.graph.r)[group %in% c('FH')] 
h2.idx <- V(sim.graph.r)[group %in% c('notFH')] 

f.idx <- V(sim.graph.r)[group %in% c('FH', 'FnotH')] 
f1.idx <- V(sim.graph.r)[group %in% c('FH')] 
f2.idx <- V(sim.graph.r)[group %in% c('FnotH')] 

this.r.df <- get.data.frame(sim.graph.r, 'vertices')

## the ratio here should be very close to tau
this.r.df %>% filter(group %in% c('FH', 'FnotH')) %>%
              summarise(y.F.H = sum(y.FH + y.notFH),
                        d.F.H = sum(d.FH + d.notFH),
                        ratio = y.F.H / d.F.H)

# on the other hand, this ratio should always be 1, no matter what the value
# of tau is, since reports from F about notH should not be affected
this.r.df %>% filter(group %in% c('FH', 'FnotH')) %>%
              summarise(y.F.notH = sum(y.FnotH + y.notFnotH),
                        d.F.notH = sum(d.FnotH + d.notFnotH),
                        ratio = y.F.notH / d.F.notH)

```

## Performance notes

I wrote two versions of the code to draw a network from the
stochastic blockmodel. The first was really slow; the second is
still slow, but is probably fast enough to use here.

```{r}
## Performance notes below:

# for N=1000, drawing one graph (first way) takes:
#  user  system elapsed
# 3.715   1.515   5.243
  
# for N=1000, drawing one graph (second / current way) takes:
#  user  system elapsed
# 0.425   0.062   0.506

# system.time(g <- draw.sbm.graph(block.sizes, pref.matrix, block.names))

# g.df <- get.data.frame(g, 'vertices')
# all(g.df$degree == g.df$nothidden + g.df$hidden)

```

